Actuator line modeling of vertical-axis turbines

### 2016-04-29 paper.tex
Replace all chapter refs with papers; resolves #14

It is therefore desirable to retain a Navier--Stokes description of the flow
field and use an actuator-type model for parameterizing the turbine loading,
which dramatically drives down computational expense by removing the need for
complicated meshes with many fine cells to resolve the boundary layers near the
blade surfaces. As shown in Chapter~\ref{chap:RVAT-baseline}, the conventional
uniform actuator disk is not a good candidate for a cross-flow turbine wake
generator, never mind the fact that it does not typically compute performance
predictions. There are then two actuator methods that use blade element theory
to compute blade loading and therefore generate a wake that more closely
resembles that of an actual turbine. These are the actuator cylinder or
swept-surface model (ASSM) and the actuator line model (ALM). The ASSM solves
for the average blade loading along its path and applies this as a constant body
force term in the momentum equation. The ALM takes a similar approach but is an
unsteady method, resolving the blade element locations in time.

@@

Both the high solidity UNH-RVAT and medium-low solidity RM2 turbines were
modeled using the ALM inside a Reynolds-averaged Navier--Stokes (RANS)
simulation, closed with the standard $k$--$\epsilon$ turbulence model. These
rotors provide diverse parameters, which helped evaluate the robustness of the
ALM. The simulations were performed inside a domain similar in size to that used
in Chapter~\ref{chap:CFD} to mimic the tow tank (3.66 m wide, 2.44 m tall, 1.52
m upstream and 2.16 m downstream), with similar boundary conditions (1 m/s
inflow, no-slip walls and bottom, and a rigid slip for the top). Simulations
were run for a total of 6 seconds, with the latter half used to calculate
performance and wake statistics. Pressure--velocity coupling for the momentum
equation was achieved using the PISO (pressure implicit splitting of operators)
method. A slice of the mesh in the $x$--$y$ plane is shown in
Figure~\ref{fig:ALM-mesh}.

@@

Since the ALM is intended to be an engineering tool when coupled with RANS, it
was assumed that information about tip speed ratio due to control details---that
is, sinusoidally oscillating $\lambda$---would not be known a priori, and was
excluded, unlike the 3-D blade-resolved cases in Chapter~\ref{chap:CFD}. Note
that a systematic investigation of the effects of sinusoidal $\lambda$ was not
undertaken, but for the UNH-RVAT a one to two percentage point increase in $C_P$
was observed when running with similar parameters as the blade-resolved
simulation.

Compared to 3-D blade-resolved RANS, the ALM can solve a standalone turbine case
on the order of CPU minutes per second of simulated time versus 1,000 CPU hours
per second---a savings of 4 orders of magnitude.

@@

Verification for sensitivity to spatial and temporal grid resolution was
performed for both the UNH-RVAT and RM2 RANS cases at their optimal tip speed
rations, the results from which are plotted in
Figure~\ref{fig:RVAT-ALM-verification} and
Figure~\ref{fig:RM2-ALM-verification}, respectively. Similar to the verification
strategy employed in Chapter~\ref{chap:CFD}, the mesh topology was kept
constant, and the resolution was scaled proportional to the number of cells in
the $x$-direction $N_x$ for the base hexahedral mesh. Both models displayed low
sensitivity to the number of time steps per revolution. Spatial grid dependence,
however, was more important.

@@

Power and drag coefficient curves are plotted for the UNH-RVAT in
Figure~\ref{fig:RVAT-ALM-perf-curves}. The ALM was successful at predicting the
performance tip speed ratios up to $\lambda_0$, which suggests that dynamic
stall was being modeled accurately, but $C_P$ was overpredicted at high
$\lambda$. This may have been caused by the omission of additional parasitic
drag sources such as roughness from exposed bolt heads located far enough from
the axis to have a large effect at high rotation rates, or an underestimation of
the blade--strut connection corner drag coefficient. In Chapter~\ref{chap:RM2}
we showed how these losses can be significant even with carefully smoothed
struts and strut-blade connections. Overprediction of performance at high tip
speed ratio could also be a consequence of the Leishman--Beddoes dynamic stall
model, which can also be seen in the Darrieus VAWT momentum model results shown
in Figure 6.70 of \cite{Para2002}.

@@

Weighted averages for the momentum recovery terms were computed identically as
they were in Chapter~\ref{chap:CFD}, and are plotted in
Figure~\ref{fig:RVAT-ALM-recovery} along with the actuator disk (AD) results
from Chapter~\ref{chap:RVAT-baseline}, 3-D blade-resolved RANS results from
Chapter~\ref{chap:CFD} and experiments. The most glaring discrepancy is the
ALM's prediction of positive cross-stream advection, which is caused by the lack
of detail in the tip vortex shedding. The total for vertical advection, however,
is close to that predicted by the 3-D blade-resolved Spalart--Allmaras model.
Levels of turbulent transport due to eddy viscosity and deceleration due to the
adverse pressure gradient are between those predicted by the 3-D blade-resolved
$k$--$\omega$ SST and SA models. Overall, however, one might expect the total
wake recovery rate to be comparable between all models except the actuator disk,
which induces negative vertical advection, very little turbulent transport, and
has a positive pressure gradient contribution. These results suggests the ALM
would be an effective tool---much better than an actuator disk---for assessing
downstream spacing of subsequent CFTs, though blade--vortex interaction for very
tightly spaced rotors may not be captured, at least on relatively coarse meshes
as used here.

@@

    \caption{Weighted average momentum recovery terms for the actuator disk (AD)
    simulation from Chapter~\ref{chap:RVAT-baseline}, UNH-RVAT actuator line
    model with a $k$--$\epsilon$ RANS closure, the two 3-D blade resolved RANS
    models described in Chapter~\ref{chap:CFD}, and the experiments in
    Chapter~\ref{chap:RVAT-baseline}.}

@@

    \caption{Weighted average momentum recovery terms at $x/D=0.93$ for the RM2
        actuator line model with a $k$--$\epsilon$ RANS closure and the experiments
        described in Chapter~\ref{chap:RM2}.}

@@

Thus, the UNH-RVAT baseline case was simulated using the Smagorinsky LES
turbulence model \cite{Smagorinsky1963}, which was the first of its kind, and
serves as a good standard for LES modeling since its behavior is well-reported
in the literature. Default Smagorinsky model coefficients were used, LES filter
width was set as the cube root of the local cell volume. The tip speed ratio was
set to oscillate sinusoidally about $\lambda_0$ with a 0.19 magnitude and the
angle of the first peak at 1.4 radians---similar to the rotation presribed in
the blade-resolved RANS simulations discussed in Chapter~\ref{chap:CFD}.

@@

Since turbinesFoam is compatible with OpenFOAM's volume of fluid (VOF)
multiphase flow solver interFoam, the ALM's effectiveness should be assessed for
predicting the generation and interaction of turbines with surface waves. The
ADV positioning system developed in Chapter~\ref{chap:exp-setup} could be
retrofitted with a wave staff to measure and validate the free surface
disturbance, which could be key to predicting the effects of placing turbines in
channels at high blockage ratio.

>>>>

It is therefore desirable to retain a Navier--Stokes description of the flow
field and use an actuator-type model for parameterizing the turbine loading,
which dramatically drives down computational expense by removing the need for
complicated meshes with many fine cells to resolve the boundary layers near the
blade surfaces. As shown in \cite{Bachant2015-JoT}, the conventional uniform
actuator disk is not a good candidate for a cross-flow turbine wake generator,
never mind the fact that it does not typically compute performance predictions.
There are then two actuator methods that use blade element theory to compute
blade loading and therefore generate a wake that more closely resembles that of
an actual turbine. These are the actuator cylinder or swept-surface model (ASSM)
and the actuator line model (ALM). The ASSM solves for the average blade loading
along its path and applies this as a constant body force term in the momentum
equation. The ALM takes a similar approach but is an unsteady method, resolving
the blade element locations in time.

@@

Both the high solidity UNH-RVAT and medium-low solidity RM2 turbines were
modeled using the ALM inside a Reynolds-averaged Navier--Stokes (RANS)
simulation, closed with the standard $k$--$\epsilon$ turbulence model. These
rotors provide diverse parameters, which helped evaluate the robustness of the
ALM. The simulations were performed inside a domain similar in size to that used
in \cite{Bachant2016-BR-CFD} to mimic the tow tank experiments (3.66 m wide,
2.44 m tall, 1.52 m upstream and 2.16 m downstream), with similar velocity
boundary conditions: 1 m/s inflow, no-slip (fixed to 1 m/s) walls and bottom,
and a rigid slip condition for the top, to approximate some effects of the tow
tank's free surface. Simulations were run for a total of 6 seconds, with the
latter half used to calculate performance and wake statistics.
Pressure--velocity coupling for the momentum equation was achieved using the
PISO (pressure implicit splitting of operators) method. A slice of the mesh in
the $x$--$y$ plane is shown in Figure~\ref{fig:ALM-mesh}.

@@

Since the ALM is intended to be an engineering tool when coupled with RANS, it
was assumed that information about tip speed ratio due to control details---that
is, sinusoidally oscillating $\lambda$---would not be known a priori, and was
excluded, unlike the 3-D blade-resolved cases reported in
\cite{Bachant2016-BR-CFD}. Note that a systematic investigation of the effects
of sinusoidal $\lambda$ was not undertaken, but for the UNH-RVAT a one to two
percentage point increase in $C_P$ was observed when running with similar
parameters as the blade-resolved simulation.

Compared to 3-D blade-resolved RANS, the ALM can solve a standalone turbine case
on the order of 0.1 CPU hours per second of simulated time versus 1,000 CPU
hours per second---a savings of 4 orders of magnitude.

@@

Verification for sensitivity to spatial and temporal grid resolution was
performed for both the UNH-RVAT and RM2 RANS cases at their optimal tip speed
rations, the results from which are plotted in
Figure~\ref{fig:RVAT-ALM-verification} and
Figure~\ref{fig:RM2-ALM-verification}, respectively. Similar to the verification
strategy employed in \cite{Bachant2016-BR-CFD}, the mesh topology was kept
constant, and the resolution was scaled proportional to the number of cells in
the $x$-direction $N_x$ for the base hexahedral mesh. Both models displayed low
sensitivity to the number of time steps per revolution. Spatial grid dependence,
however, was more important.

@@

Power and drag coefficient curves are plotted for the UNH-RVAT in
Figure~\ref{fig:RVAT-ALM-perf-curves}. The ALM was successful at predicting the
performance tip speed ratios up to $\lambda_0$, which suggests that dynamic
stall was being modeled accurately, but $C_P$ was overpredicted at high
$\lambda$. This may have been caused by the omission of additional parasitic
drag sources such as roughness from exposed bolt heads located far enough from
the axis to have a large effect at high rotation rates, or an underestimation of
the blade--strut connection corner drag coefficient. In
\cite{Rawlings2008,Bachant2016-RM2-paper} it was shown how these losses can be
significant even with carefully smoothed struts and strut-blade connections.
Overprediction of performance at high tip speed ratio could also be a
consequence of the Leishman--Beddoes dynamic stall model, which can also be seen
in the Darrieus VAWT momentum model results shown in Figure 6.70 of
\cite{Para2002}.

@@

Weighted averages for the terms in the streamwise momentum equation were
computed identically as they were in \cite{Bachant2015-JoT,Bachant2016-BR-CFD},
and are plotted in Figure~\ref{fig:RVAT-ALM-recovery} along with the actuator
disk (AD) results from \cite{Bachant2015-JoT}, 3-D blade-resolved RANS results
from \cite{Bachant2016-BR-CFD} and experiments. The most glaring discrepancy is
the ALM's prediction of positive cross-stream advection, which is caused by the
lack of detail in the tip vortex shedding. The total for vertical advection,
however, is close to that predicted by the 3-D blade-resolved Spalart--Allmaras
model. Levels of turbulent transport due to eddy viscosity and deceleration due
to the adverse pressure gradient are between those predicted by the 3-D
blade-resolved $k$--$\omega$ SST and SA models. Overall, however, one might
expect the total wake recovery rate to be comparable between all models except
the actuator disk, which induces negative vertical advection, very little
turbulent transport, and has a positive pressure gradient contribution. These
results suggests the ALM would be an effective tool---much better than an
actuator disk---for assessing downstream spacing of subsequent CFTs, though
blade--vortex interaction for very tightly spaced rotors may not be captured, at
least on relatively coarse meshes as used here.

@@

    \caption{Weighted average momentum recovery terms for the actuator disk (AD)
        simulation from \cite{Bachant2015-JoT}, UNH-RVAT actuator line model with a
        $k$--$\epsilon$ RANS closure, the two 3-D blade resolved RANS models
        described in \cite{Bachant2016-BR-CFD}, and the experiments reported in
        \cite{Bachant2015-JoT}.}

@@

    \caption{Weighted average momentum recovery terms at $x/D=0.93$ for the RM2
        actuator line model with a $k$--$\epsilon$ RANS closure and the experiments
        described in \cite{Bachant2016-RM2-paper}.}

@@

Thus, the UNH-RVAT baseline case was simulated using the Smagorinsky LES
turbulence model \cite{Smagorinsky1963}, which was the first of its kind, and
serves as a good standard for LES modeling since its behavior is well-reported
in the literature. Default Smagorinsky model coefficients were used, LES filter
width was set as the cube root of the local cell volume. The tip speed ratio was
set to oscillate sinusoidally about $\lambda_0$ with a 0.19 magnitude and the
angle of the first peak at 1.4 radians---similar to the rotation presribed in
the blade-resolved RANS simulations discussed in \cite{Bachant2016-BR-CFD}.

@@

Since turbinesFoam is compatible with OpenFOAM's volume of fluid (VOF)
multiphase flow solver interFoam, the ALM's effectiveness should be assessed for
predicting the generation and interaction of turbines with surface waves. Tow
tank experiments could be performed with a wave gauge to measure and validate
the free surface disturbance, which could be key to predicting the effects of
placing turbines in channels at high blockage ratio.
